# Adapting tidy_cpue_index for use with hook and line data in addition to trawl.

Extracting raw data
```{r}
trawl <- gfdata::get_cpue_index()
hl <- gfdata::get_cpue_index(gear = "hook and line")


dat <- trawl
dat2 <- hl
```

- Changed min_positive_tows to min_positive_fe for consistency throughout the
code. Applied this throughout gfplot, gfsynopsis and cpue_reports. Not sure if
there are other packages to adapt.
- Separated out filters that apply to both trawl and hl from gear-type-specific
filters
- Changed "hours fished" to "effort"
- Moved creation of alt_year to within the if use_alt_year statement to save 
computation time
- Removed/comemnted out code which was either redundant or did nothing
- 

Testing function
```{r}
tidy <- tidy_cpue_index(trawl, "yelloweye rockfish")
tidy2 <- tidy_cpue_index(hl, "yelloweye rockfish")
```

Set arguments in order to step through function:
```{r}
species_common = "yelloweye rockfish"
year_range = c(1996, as.numeric(format(Sys.Date(), "%Y")) - 1)
alt_year_start_date = "04-01"
use_alt_year = FALSE
lat_range = c(48, Inf)
min_positive_fe = 100
min_positive_trips = 5
min_yrs_with_trips = 5
area_grep_pattern = "^3C|^3D|^5A|^5B|^5C|^5D|^5E"
lat_band_width = 0.1
depth_band_width = 25
clean_bins = TRUE
depth_bin_quantiles = c(0.001, 0.999)
min_bin_prop = 0.001
lat_bin_quantiles = c(0, 1)
gear = c("bottom trawl", "hook and line")
```

```{r}
glimpse(tidy2)
```

How many rows of data are lost for hook and line data for those removed due
to NA's?
Total rows:
```{r}
nrow(dat)
```
1982888

Count of NA's for various columns that may be filtered on:
```{r}
dat_short <- dat %>% 
  select(vessel_name, vessel_registration_number, fe_start_date, fe_end_date, best_depth, latitude, longitude)
dat_short <- dat_short[!duplicated(dat_short, , drop = FALSE)]
colSums(is.na(dat_short))
```
vessel_name   vessel_registration_number   fe_start_date  fe_end_date   best_depth  latitude    longitude 
    8857                4497               495052         495050        198928      196644        196670    
    
We lose about 10% of records (note that these numbers are species catch records,
not individual fe's - there are about 230,000 unique fe_id's)
by eliminating records with no lat/long or depth,
25% of records by filtering out those with no fe start or end date,
0.25% for those with no vessel reg numer.

Filtering out records without fe dates has been moved to just trawl data
filtering in the function since it is dependent on dates for effort
calculation.


# Hooks per fe for hook and line effort

Longline gear specs are different in GFFOS than GFBio, so code from IPHC data
extraction for effort cannot be used. Many hook data are missing, on the order
of 350,000 out of 500,000 records in the gear specs table, out of 
approximately 2,000,000 records
