---
output:
  md_document:
    variant: gfm
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# gfplot: An R package for data extraction and plotting of British Columbia groundfish data

[![Travis-CI Build Status](https://travis-ci.org/pbs-assess/gfplot.svg?branch=master)](https://travis-ci.org/pbs-assess/gfplot)
<!-- [![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip) -->

Facilitates the creation of an annual groundfish data synopsis report with [gfsynopsis](https://github.com/pbs-assess/gfsynopsis).

Note that the documentation is incomplete in some places. Please post in the [issue tracker](https://github.com/pbs-assess/gfplot/issues) if you have questions or suggestions on how the package or its documentation could be improved. We welcome pull requests!

In addition to the help available through the R console, a [web version of the documentation is available](https://pbs-assess.github.io/gfplot/index.html).

# Installation

The gfplot package can be installed and loaded with:

```{r, eval=FALSE}
install.packages("INLA", repos = c(getOption("repos"), 
  INLA = "https://inla.r-inla-download.org/R/stable"), dep = TRUE)
# install.packages("devtools")
devtools::install_github("pbs-assess/gfplot")
```

```{r}
library(gfplot)
```

Functions specific to PBS:

```{r}
fns <- ls("package:gfplot")
sort(fns[grepl("get", fns)])
```

Generic functions for any similarly formatted data:

```{r}
sort(fns[grepl("tidy", fns)])
```

```{r}
sort(fns[grepl("fit", fns)])
```

```{r}
sort(fns[grepl("plot", fns)])
```
