---
output:
  md_document:
    variant: markdown_github
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# gfplot: An R package for data extraction (at PBS) and plotting (of any) groundfish data

[![Travis-CI Build Status](https://travis-ci.org/seananderson/gfplot.svg?branch=master)](https://travis-ci.org/seananderson/gfplot)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/seananderson/gfplot?branch=master&svg=true)](https://ci.appveyor.com/project/seananderson/gfplot)
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)

Facilitates the creation of an annual groundfish data synopsis report with [gfsynopsis](https://github.com/seananderson/gfsynopsis).

# Installation

The gfplot package is *not* ready for use yet. In particular, the documentation is far from complete. However, it can be installed and loaded with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("seananderson/gfplot")
```

```{r}
library("gfplot")
```

Functions specific to PBS:

```{r}
fns <- ls("package:gfplot")
sort(fns[grepl("get", fns)])
```

Generic functions for any similarly formatted data:

```{r}
sort(fns[grepl("tidy", fns)])
```

```{r}
sort(fns[grepl("fit", fns)])
```

```{r}
sort(fns[grepl("plot", fns)])
```
