---
title: "Exploring CPUE for five species of flatfish"
author: "Sean Anderson and Elise Keppel"
output: html_document
---


```{r, knitr-opts, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.asp = 0.618,
  autodep = TRUE,
  cache = TRUE,
  cache.comments = FALSE
)
```


```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(gfplot)
```

Extract catch and effort data from all databases for species of interest:

```{r, eval=FALSE}
species <- c("Arrowtooth Flounder", "English Sole", "Dover Sole", "Southern Rock Sole",
  "Petrale Sole") 
d <- gfplot::get_cpue_historic(species, fishing_year = FALSE, end_year = 2017)
saveRDS(d, file = "../data/flatfish-cpue-data.rds", compress = FALSE)
```

Read the data in, clean up the column names, and create columns for the major statistical areas:

```{r}
d <- readRDS("data/historic-cpue-data.rds")
```

Filter down to remove all rows of data with missing hours fished, missing total catch, or when the total catch or hours fished are 0. 
```{r}
d <- d %>%
  filter(!is.na(hours_fished), !is.na(total), total > 0, hours_fished > 0) %>%
  filter(!is.na(fyear))
```

Now let's calculate the arithmetic and geometric mean CPUE for each species in each year:

```{r}
d_sum <- d %>%
  group_by(area, fyear, species_common_name) %>%
  summarise(
    catch = sum(total, na.rm = TRUE),
    sum_hours_fished = sum(hours_fished, na.rm = TRUE),
    arith_cpue = sum(total, na.rm = TRUE) / sum(hours_fished, na.rm = TRUE),
    geo_cpue = exp(mean(log(total / hours_fished), na.rm = TRUE))
  ) %>%
  ungroup()
```

Let's create a version that is scaled by the geometric mean for plotting to mimic past research documents:

```{r}
d_scaled <- d_sum %>%
  mutate(arith_cpue = arith_cpue / exp(mean(log(arith_cpue)))) %>%
  mutate(geo_cpue = geo_cpue / exp(mean(log(geo_cpue))))
```

Plot of the scaled version with locality/all in different columns:

```{r}
d_scaled %>%
  tidyr::gather(cpue_type, cpue_value, arith_cpue:geo_cpue) %>%
  ggplot(aes(fyear, cpue_value, linetype = cpue_type)) +
  geom_line() +
  geom_vline(xintercept = 1996, lty = 3) +
  facet_grid(area~species_common_name, scales = "free_y") +
  xlab("") +
  ylab("CPUE divided by geometric mean") +
  ylim(0, NA) +
  theme_pbs() +
  labs(linetype = "CPUE type")
```

Plot of the scaled version with CPUE type in different columns:

```{r}
d_scaled %>%
  tidyr::gather(cpue_type, cpue_value, arith_cpue:geo_cpue) %>%
  ggplot(aes(fyear, cpue_value, colour = species_common_name)) +
  geom_line() +
  geom_vline(xintercept = 1996, lty = 3) +
  facet_grid(area~cpue_type, scales = "free_y") +
  xlab("") +
  ylab("CPUE divided by geometric mean") +
  ylim(0, NA) +
  theme_pbs() +
  labs(colour = "Species")
```

Save the unstandardized and unscaled data:

```{r}
saveRDS(d_sum, file = "../data/flatfish-annual-cpue.rds", compress = FALSE)
```
