---
title: "Split catch by maturity"
author: "Philina English"
date: "05/05/2021"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  autodep = TRUE,
  fig.path = "pcod-knitr-figs/",
  cache.path = "pcod-knitr-cache/"
)
```

# Setup

If you don't already have the package installed, then run:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("seananderson/gfplot")
```

First we will load the package along with gfdata, ggplot2, and dplyr since we will use them within our code later.

```{r, cache=FALSE}
library(gfplot)
library(gfdata)
library(ggplot2)
library(dplyr)
options(scipen=999)
```


# Get the synoptic survey set and biological sample data 

As an example, we could extract the data with the following gfdata function calls if we were on a DFO laptop, with appropriate database permissions, and on the PBS network:


```{r, eval=FALSE}
d_survey_sets  <- get_survey_sets("pacific cod")
d_survey_samples  <- get_survey_samples("pacific cod")
```

In addition to the individual `get_*()` functions, there is a function `cache_pbs_data()` that runs all the `get_*()` functions and caches the data in a folder that you specify. I'm going to use this approach here because I'm not currently on the PBS network and it saves running the SQL queries each time the R Markdown document is recompiled.

The helper function `cache_pbs_data()` will extract all of the data into a series of `.rds` files into whatever folder you specify to the `path` argument. I'll wrap it in a quick check just to make sure we don't download the data twice if we build this document again.

```{r}
if (!file.exists(file.path("pcod-cache", "pbs-survey-sets.rds"))) { # quick check for speed
  cache_pbs_data("pacific cod", path = "pcod-cache")
}
```

Let's read those data files in to work with here.

```{r}
cache             <- file.path("pcod-cache")
d_survey_sets     <- readRDS(file.path(cache, "pbs-survey-sets.rds"))
d_survey_samples  <- readRDS(file.path(cache, "pbs-survey-samples.rds"))
```

Now let's create a folder to hold some PDF versions of our figures:

```{r}
figs <- file.path("pcod-figs")
dir.create(figs, showWarnings = FALSE)
```


```{r}
d_split <- split_catch_maturity(d_survey_sets, d_survey_samples, 
  survey = c("SYN HS", "SYN QCS"),
  years = NULL, 
  sample_id_re = FALSE, 
  year_re = FALSE,
  p_threshold = 0.5,
  plot = TRUE
)
print(d_split$maturity_plot)
```


Underlying biological data and model details are saved in object m

```{r}
d_split$m$model
```

Both year and sample id as random effects
```{r}
d_split_re <- split_catch_maturity(d_survey_sets, d_survey_samples, 
  survey = c("SYN HS", "SYN QCS"),
  years = NULL, 
  sample_id_re = TRUE, 
  year_re = TRUE,
  p_threshold = 0.5,
  plot = TRUE
)
print(d_split_re$maturity_plot)
```

```{r}
d_split_re$m$model
```


Just year as a random effect
```{r}
d_split_yr <- split_catch_maturity(d_survey_sets, d_survey_samples, 
  survey = c("SYN HS", "SYN QCS"),
  years = NULL, 
  sample_id_re = FALSE, 
  year_re = TRUE,
  p_threshold = 0.5,
  plot = TRUE
)

print(d_split_yr$maturity_plot)
```

While the above plot is saved by the splitting function, it can also be rebuilt and modified using the plot functions applied to the saved model object named 'm'.

```{r}
plot_mat_ogive(d_split_yr$m)
```


For models with year as a random effect, there's a special function for plotting annual ogives in colour.
```{r}
plot_mat_annual_ogives(d_split_yr$m)
```

```{r}
plot_mat_annual_ogives(d_split_re$m)
```


```{r}
d_split_sa <- split_catch_maturity(d_survey_sets, d_survey_samples, 
  survey = c("SYN HS", "SYN QCS"),
  years = NULL, 
  sample_id_re = TRUE, 
  year_re = FALSE,
  p_threshold = 0.5,
  plot = TRUE
)
print(d_split_sa$maturity_plot)
```


The "data" element contains the original d_survey_sets dataframe with the addition of:

est_sample_mass - estimated total mass of individually sampled fish
measured_weight - measured total mass of individually sampled fish
n_weights - count of individually weighed fish
n_mature - number of mature fish sampled
n_sampled - total fish individually sampled fish

mass_ratio_mature - ratio of mass that represents mature fish

raw_error - catch minus samples in kg; < 0 should be impossible
perc_sampled - percent by mass of catch sampled; >1 suggests either fish were smaller than predicted from lengths, or a discrepency in data entry  

split_dens_type - what are the units of the following density columns
adult_density - estimated biomass density of mature fish
imm_density - estimated biomass density of immature fish


```{r}
glimpse(d_split$data)
```

```{r}
ggplot(data = d_split$data, aes(catch_weight, perc_sampled)) +
  geom_point(alpha = 0.25) +
  # ylim(0,10) + xlim(0, 70) +
  theme_classic()
```

There are some samples that have higher mass of individual fish samples than the reported total catch

```{r}
ggplot(data = d_split$data, aes(catch_weight, est_sample_mass/1000, colour = n_sampled-n_weights)) +
  geom_point(alpha = 0.5) +
  xlim(0, 100) + ylim(0,100) +
  theme_classic()
```

```{r}
ggplot(data = d_split$data, aes(unsampled_catch)) +
  geom_histogram(alpha = 0.5) +
  # xlim(-20, 100) +
  scale_x_log10() +
  theme_classic()
```

```{r}
ggplot(data = d_split$data, aes(catch_weight, unsampled_catch, colour = n_sampled-n_weights)) +
  geom_point(alpha = 0.5) +
  xlim(0, 250) + ylim(-20, 0) +
  theme_classic()
```

```{r}
print(d_split$mass_plot)
```
